A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-15 tex/codatadp.tex
played with headings
More reorg, prove model construction produces values.

For a $\mu$-term $t_1$ with subterm $t_2$, 
the \emph{interpretation of $t_2$ in $t_1$} is the $\mu$-term $t_1\interp{t_2}^\emptyset$ as returned by the following recursive procedure,
where $x \mapsto \mu x. \const{C}( \vec u ) \in t_1$ indicates that $\mu x. \const{C}( \vec u )$ is the subterm of $t_1$ that binds this occurrence of variable $x$.
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
t_1\interp{x}^M & = & \mu x. \const{C}( t_1\interp{ \vec u}^{M \cup \{ x \}}) & x \not\in M, x \mapsto \mu x. \const{C}( \vec u ) \in t_1  \\
t_1\interp{x}^M & = & x & \text{otherwise} \\
t_1\interp{\mu x. \const{C}( \vec u )}^M & = & \mu x. \const{C}( t_1\interp{ \vec u }^{M \cup \{ x \}} ) & x \not\in M \\
t_1\interp{\mu x. \const{C}( \vec u )}^M & = & x & \text{otherwise}
\end{array}\]
We will often write $t_1\interp{t_2}$ as shorthand for $t_1\interp{t_2}^\emptyset$.
We will write $\mathcal{T}( t_1 )$ to denote the set consisting of $t_1\interp{t_2}$ for all subterms $t_2$ of $t_1$,
and write $\mathcal{T}( S )$ to denote the union of $\mathcal{T}( t )$ for each $t \in S$ where $S$ is a set of terms.

@@

We say that $t \vsimv{x} s$ if and only if $\mu$-terms $t$ and $s$ are syntactically equivalent
for some renaming that avoids capturing any variable other than $x$.
For instance, 
$\mu x. \const{D}( x ) \vsimv{y} \mu x. \const{D}( y )$,
$\mu x. \const{C}( x, x ) \vsimv{x} \mu y. \const{C}( x, y )$ and
$\mu x. \const{C}( z, x ) \vsimv{z} \mu y. \const{C}( z, y )$,
but
$\mu x. \const{D}( x ) \not\vsimv{x} \mu x. \const{D}( y )$ and
$\mu x. \const{C}( x, x ) \not\vsimv{y} \mu y. \const{C}( x, y )$.

%\begin{lemma}
%If $t_1$ and $u$ are normal, and if $u \not\vsimv{x} t_1\interp{t_2}$ for all $t_2 \in \mathcal{T}( t_1 )$,
%then $t_1 \{ x \mapsto u \}$ is normal.
%\end{lemma}
%\begin{proof}
%\rem{TODO}
%\end{proof}

@@

In the following, we consider only normal models.
Thus, when constructing a model $\M$ for $\Ec$,
it remains only to define how wrong-applied selector terms and variables in $\Ec$ are interpreted.
For the latter, this will be based on the mapping $\Val$ constructed in step two of our calculus.
Given a normal model $\M$, we compute the \emph{completion} of $\Val$, which we will write as $\Val^\ast$,
by exhaustively applying the following update to $\Val$:

\(
\inferrule{
  x : \tau \in FV( \Val ) 
  \quad
  \mu x. t \sim v, v \in \M( \tau )
  \quad
  (\mu x. t) \not\sim_x u \text{ for all } u \in \mathcal{T}(\vrange(\Val))
}{
  \Val := \Val \{ x \mapsto \mu x. t \}
}
\)

Since this rule removes one variable from the set $FV( \Val )$ and does not add any additional variables to $FV( \Val )$, 
it can only be applied a finite number of times.
We then replace all values in the range of $\Val^\ast$ by their normal forms.
We first state the following properties of $\Val^\ast$, 
where we write $\muvar( t )$ to denote $t$ if $t$ is a variable, or $x$ if $t$ is of the form $\mu x. u$.
%which amounts to saying that 
%equivalence class are interpreted as values,
%the interpretation of an equivalence class is not sensitive to the context in which it is interpreted,
%and that each equivalence classes has a distinct interpretation.

\begin{lemma}
\label{lem:mu-norm-arg}
If $\vec u$ are normal, then if $t = \nf{(\mu x. \const{C}( \vec u ) )} = \mu x. \const{C}( \vec w )$,
then $\vec u \vsim t\interp{\vec w}$.
\end{lemma}

\begin{lemma}
\label{lem:mu-cong}
If $t_u = \mu x. \const{C}( \vec u )$ and $t_w = \mu y. \const{C}( \vec w )$ are normal,
then $t_u\interp{ \vec u } \vsim t_w\interp{ \vec w }$ if and only if $t_u \vsim t_w$.
\end{lemma}

\begin{lemma} 
\label{lem:model-completion}
The following are true of $\Val^\ast$:
\begin{enumerate}
\item $\Val^\ast \ec{x} \in \M( \tau )$ for all $\ec{x} : \tau$ of $\Ec$,
\item
$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
for all $t_1\interp{t_2} \in \mathcal{T}(\Val)$ where $\muvar(t_2)=v_{\ec{x}}$,
\item 
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we must argue that $FV( \Val^\ast ) = \emptyset$, and that $\Val^\ast \ec{x}$ is not cyclic for
$\ec{x} : \tau$ when $\tau \not\in \Codata$.
\rem{TODO}
\qed
\end{proof}

@@

If $t$ is a selector term of the form $s^j_k( u )$,
since \rn{Split} does not apply, 
$\ec{u}$ must contain a term of the form $C_i( s^1_i( u ), \ldots, s^{n}_i( u ) )$ for some $i$.
By our construction and since \rn{Unify$_1$} and \rn{Unify$_2$} do not apply to $\Fc$, 
$\Val^\ast \ec{ u }$ must be of the form $t_u = \mu v_{\ec{u}}. C_i( u_1, \ldots, u_n )$,
where $\muvar( u_1 ) = v_{\ec{s^1_i(u)}}$, $\ldots$, $\muvar(u_n) = v_{\ec{s^n_i( u )}}$.
Moreover, $\M( u )$ is of the form $t_w = \mu x. C_i( w_1, \ldots, w_n )$,
which by the induction hypothesis is bisimilar with $\Val^\ast \ec{u}$.
If $k = i$, then $\M( t )$ is bisimilar with $t_w\interp{w_j}$,
which by Lemma~\ref{lem:mu-cong} is bisimilar with $t_u\interp{u_j}$,
which by Lemma~\ref{lem:model-completion}.2 is equal to $\Val^\ast \ec{ s^j_k( u )}$.
Thus, $\M( t ) \vsim \Val^\ast \ec{ s^j_k( u )} = \Val^\ast \ec{t}$.
If $k \neq i$, then we interpret $s^j_k( \M( u ) )$ as the value in $\M( \tau )$ that is bisimilar with $\Val^\ast \ec{ t }$.
Since \rn{Cong} does not apply and due to Lemma~\ref{lem:model-completion}.3, 
pairs of wrongly-applied selectors are assigned the same value if and only if they reside in the same equivalence class,
and thus this definition is well-defined.

@@

By Lemmas~\ref{lem:t},~\ref{lem:rs}, and~\ref{lem:ss}, the calculus described in this section is sound and complete for $\thD$.

>>>>

For a $\mu$-term $t_1$ with subterm $t_2$, 
the \emph{interpretation of $t_2$ in $t_1$} is the $\mu$-term $t_1\interp{t_2}^\emptyset$ as returned by the following recursive procedure,
where $x \mapsto \mu x. \const{C}( \vec u ) \in t_1$ indicates that $\mu x. \const{C}( \vec u )$ is the subterm of $t_1$ that binds this occurrence of variable $x$.
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
t_1\interp{x}^M & = & \mu x. \const{C}( t_1\interp{ \vec u}^{M \cup \{ x \}}) & x \not\in M, x \mapsto \mu x. \const{C}( \vec u ) \in t_1  \\
t_1\interp{x}^M & = & x & \text{otherwise} \\
t_1\interp{\mu x. \const{C}( \vec u )}^M & = & \mu x. \const{C}( t_1\interp{ \vec u }^{M \cup \{ x \}} ) & x \not\in M \\
t_1\interp{\mu x. \const{C}( \vec u )}^M & = & x & \text{otherwise}
\end{array}\]
We will often write $t_1\interp{t_2}$ as shorthand for $t_1\interp{t_2}^\emptyset$.

@@

\begin{lemma}
\label{lem:mu-norm-arg}
If $\vec u$ are normal, then if $t = \nf{(\mu x. \const{C}( \vec u ) )} = \mu x. \const{C}( \vec w )$,
then $\vec u \vsim t\interp{\vec w}$.
\end{lemma}
\begin{proof}
%Since $u_i$ is normal, $w_i$ is the result of replacing all subterms $v$ of $u_i$ such that $t\interp{v} \sim t$ with $x$.
\rem{TODO}
\end{proof}

\begin{lemma}
\label{lem:mu-cong}
If $t_u = \mu x. \const{C}( \vec u )$ and $t_w = \mu y. \const{C}( \vec w )$ are normal,
then $t_u\interp{ \vec u } \vsim t_w\interp{ \vec w }$ if and only if $t_u \vsim t_w$.
\end{lemma}
\begin{proof}
\rem{TODO}
\end{proof}

@@

In the following, we consider only normal models.
Thus, when constructing a model $\M$ for $\Ec$,
it remains only to define how wrong-applied selector terms and variables in $\Ec$ are interpreted.
For the latter, this will be based on the mapping $\Val$ constructed in step two of our calculus.
We compute the \emph{completion} of $\Val$ for normal model $\M$, which we will write as $\Val^\ast$,
which we construct by assigning values from $\M$ to unassigned variables in the domain of $\Val$.

First, we need the following definition.
We say that $t \vsimv{x} s$ if and only if $\mu$-terms $t$ and $s$ are syntactically equivalent
for some renaming that avoids capturing any variable other than $x$.
For instance, 
$\mu x. \const{D}( x ) \vsimv{y} \mu x. \const{D}( y )$,
$\mu x. \const{C}( x, x ) \vsimv{x} \mu y. \const{C}( x, y )$ and
$\mu x. \const{C}( z, x ) \vsimv{z} \mu y. \const{C}( z, y )$,
but
$\mu x. \const{D}( x ) \not\vsimv{x} \mu x. \const{D}( y )$ and
$\mu x. \const{C}( x, x ) \not\vsimv{y} \mu y. \const{C}( x, y )$.
For a variable $x$ of type $\tau$ and a fixed normal model $\M$,
we write $\mathcal{T}^\M_x( \Val )$ to denote the set consisting of all values $v$ from $\M( \tau )$
such that $v \vsim_x t_1\interp{t_2}$ for some term $t_1$ with subterm $t_2$ occurring in the range of $\Val$.
Our completion $\Val^\ast$ is constructed by exhaustively applying the following update to $\Val$:

\(
\inferrule{
  x : \tau \in FV( \Val ) 
  \quad
  (\mu x. t) \sim v
  \quad
  v \in \M( \tau )
  \quad
  v \not\in \mathcal{T}^\M_x(\Val)
}{
  \Val := \Val \{ x \mapsto \mu x. t \}
}
\)

In this update, we choose an unassigned variable in $\Val$
and assign it a value that is not bisimilar to one occurring in $\mathcal{T}^\M_x(\Val)$.
Since this update removes one variable from the set $FV( \Val )$ and does not add any additional variables to $FV( \Val )$, 
it can only be applied a finite number of times.
We then replace $\Val^\ast \ec{x}$ with $\nf{(\Val^\ast \ec{x})}$ for all $\ec{x}$.
We first state the following properties of $\Val^\ast$, 
where we write $\muvar( t )$ to denote $t$ if $t$ is a variable, or $x$ if $t$ is of the form $\mu x. u$.

\begin{lemma} 
\label{lem:model-completion}
If $\Val$ is constructed for a saturated $\Ec$,
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, then:
\begin{enumerate}
\item $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$, for all $\ec{x} : \tau$ of $\Ec$,
\item
$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
for all $t_1$ with subterm $t_2$ in the range of $\Val^\ast$ where $\muvar(t_2)=v_{\ec{x}}$,
\item 
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable $v_{\ec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else $v_{\ec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one (call the value in its domain $c$).
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign $c$ to $v_{\ec{x}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction, 
since there are only a finite number of closed terms in $\mathcal{T}^\M_x(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_x(\Val^\ast)$ to $v_{\ec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Cyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form 
$\{ y \mapsto t \}$, where $t$ is acyclic when the type of $y$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$.

\rem{TODO}
\qed
\end{proof}

@@

If $t$ is a selector term of the form $s^j_k( u )$,
since \rn{Split} does not apply, 
$\ec{u}$ must contain a term of the form $C_i( s^1_i( u ), \ldots, s^{n}_i( u ) )$ for some $i$.
By our construction of $\Val^\ast$ and since \rn{Unify$_1$} and \rn{Unify$_2$} do not apply to $\Fc$, 
$\Val^\ast \ec{ u }$ must be of the form $t_u = \mu v_{\ec{u}}. C_i( u_1, \ldots, u_n )$,
where $\muvar( u_1 ) = v_{\ec{s^1_i(u)}}$, $\ldots$, $\muvar(u_n) = v_{\ec{s^n_i( u )}}$,
and $\M( u )$ is of the form $t_w = \mu x. C_i( w_1, \ldots, w_n )$,
which by the induction hypothesis is bisimilar with $\Val^\ast \ec{u}$.
If $k = i$, then $\M( t )$ is bisimilar with $t_w\interp{w_j}$,
which since $t_w \vsim t_u$ by Lemma~\ref{lem:mu-cong} is bisimilar with $t_u\interp{u_j}$,
which by Lemma~\ref{lem:model-completion}.2 is equal to $\Val^\ast \ec{ s^j_k( u )}$.
Thus, $\M( t ) \vsim \Val^\ast \ec{ s^j_k( u )} = \Val^\ast \ec{t}$.
If $k \neq i$, then we interpret $s^j_k( \M( u ) )$ as the value in $\M( \tau )$ that is bisimilar with $\Val^\ast \ec{ t }$.
Since \rn{Cong} does not apply and due to Lemma~\ref{lem:model-completion}.3, 
pairs of wrongly-applied selectors are assigned the same value if and only if they reside in the same equivalence class,
and thus this definition is well-defined.

@@

By Lemmas~\ref{lem:t},~\ref{lem:rs}, and~\ref{lem:ss}, the calculus is sound and complete for $\thD$.
